\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-02 chapters/bci.tex
New patient table
Comments hakim

\todo{comments liesel paradigms})
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}

People suffering from severe paralysis or disability due to
neurodegenenerative disease, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

@@

The key problem for severely paralyzed patients is that the mind wants to go
where the body can't follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for
assistive technology and communication devices.
},
a system that directly couples actions in the `real world' to the mental state of
the user.

Starting in the 1970's from methods to establish basic communication from
minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab
setting and can be used as assistive communication technology by patients with
impaired speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting edge examples of experimental use include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

\section{A direct interface to the brain}

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what's going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to establish the
activity of a specific brain region
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical activity are be faster and more
informative.

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
For this reason, a \ac{bci} operates using a defined \emph{paradigm}, the manner in
which interaction is performed.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the inteded letter A or B
in Figure~\ref{fig:bci/loop}.
The \ac{bci} paradigm comprises, if applicable, the manner of stimulation and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

@@

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.
As touched upon earlier,the \emph{invasiveness} of the technology forms an
important concept in determining its suitability for a given application and
patient.
As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.

@@

\section{\Ac{bci} paradigms}
\label{sec:bci/paradigms}

@@

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as \emph{passive
\ac{bci}}.
From the user's point of view, this concept is probably the most attractive,
since their intentions or cognitive state state would be inferred from just
their brain activity.
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
Furthermore, active paradigms\todo{"active" is used in a fuzzy way: active BCI is self-paced, reactive is in response to a cued event. Further down that becomes more clear.
The "active participation": is needed in both cases as shown in Fig. 1.3.
I suggest to say active participation} make it easier to quickly collect training data for
supervized machine learning since the conditions can be more easily controlled
by the user or the stimulation.

Paradigms with high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{%
  \cite{Becker2022,Thompson2019}
}.
\todo{footnote on how this just means it needs to be more adapted,
inclusivity, harmful to progress, paraphrase abstracts of articles}
\todo{user training more than with reactive, task instruction}

Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms usually require higher effort
and can induce fatigue due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

\section{A case study: the visual oddball \ac{bci}}

>>>>

\todo{comments liesel paradigms})
\todo{spellcheck}
\todo{are all acronyms defined}
\todo{fix all latex warnings to make sure no refs are missing (or grep on that
specific warning)}
\todo{uniform ref or autoref}
\todo{don't capitalize figure, table, algorithm,...}
\todo{define short captions for tables and figures}
\todo{fix lacheck/chktex notifications}
\todo{use terminology individuals with severe
speech and physical impairments (SSPI), avoid phrasing 'suffering from'}
\todohsm{more subtitles}
\todofc{double check statements about what has been published 'for the sake of clarity, the introduction has been mofed to"}
\todofc{chapters should look more like full papers}

People with severe physical impairment due to
neurodegenenerative disease, stroke, traumatic brain or spine injury
become entirely reliant on caregivers or family members in their day to day
routine.
Wouldn't it be great if we could help them regain some quality of life by
making them more independent?

@@

The key problem for severely paralyzed patients is that the mind wants to go
where the body cannot follow.
Couldn't we therefore design a system that directly interacts with the mind?
This system can then form an interface between the mind and any kind of
technology, like a robot arm or speech synthesizer, interacting with the world
on the user's behalf.
Such a device is the \ac{bci}\footnote{Sometimes also
termed brain-machine interface (BMI) when coupled to a physical actuator, like
a robot arm or an exoskeleton.
The term \ac{bci} is usually preferred for
assistive technology and communication devices.
},
a system that directly couples actions in the `real world' to the mental state of
the user.

Starting in the 1970's from methods to establish basic communication from
minimal brain activity (yes/no, left/right, \ldots)~\cite{Wolpaw2002}, \acp{bci} have gradually evolved
into sophisticated interaction schemes that restore or enhance many capabilities
by bypassing defective parts of the nervous or muscle system.
By now, classic \acp{bci} methods are starting to mature outside the lab
setting and can be used as assistive communication technology by patients with
impaired speech~\cite{Wolpaw2018,Milekovic2018}.
Impressive cutting edge experimental examples include a brain-spine
interface allowing a paraplegic patient to walk again\cite{Lorach2023},
a \ac{bci} translating internal speech to a facial avatar speaking the imagined
text~\cite{Metzger2023}, and fast communication through decoding imagined
handwritten symbols~\cite{Willett2021}.
Recently, the idea has also gained popular traction through Elon Musk's
NeuraLink~\cite{Musk2019} initiative, which imagines a multipurpose \ac{bci}.

@@

\section{A direct interface to the brain}
\todohsm{Start with state of the art before theses.}
\todohsm{objectives at the end of chapter 1}
\todohsm{Introduction part: chapter 1 rationale, chapter 2: literature/sota}
\todohsm{make less verbose}

@@

Let's break this definition down and focus on its separate parts.
First of all, we need to identify a signal that is a direct representation of
what is going on in the brain.
Section~\ref{sec:bci-recording} gives an overview of the available options.
One might for instance measure the fluctuations in bloodflow to establish the
activity of a specific brain region
However, this signal reacts too slowly to reflect the real-time activity that is
of interest for our purposes, and it carries little information other than the
brain regions where it is originating.
A better candidate is the neuronal electrical activity.
The brain contains 86 billion neurons, which are highly interconnected cells that are
the smallest units of information processing.
The \emph{action potential} and its related \emph{postsynaptic potential}
is an electrical pulse occurring when a neuron receives
input from another neuron and is activated.
The firing of a neuron, or the combined firing of groups of
synchronized neurons, thus generate an electrical field in and near the brain,
which can be measured using electrodes.
Fluctuations in the brain's electrical activity are be faster and more
informative.

The recorded brain waves form only one part of the interaction scheme presented
in Figure~\ref{fig:bci/loop}.
It would make little sense to start digging around at random in all the brain
activity to try to extract the desired output.
It would help greatly if we knew what we are looking for.
A \ac{bci} operates according to a specified paradigm, which defines the method by which interaction is conducted.
This usually means that the user is instructed to perform a specific task, like
attending a specific flickering stimulus.
We can now exploit background knowledge from prior neuroscientific research
that the brain response evoked by attended stimuli is different of those
unattended,
and determine which one of a set of targets was attended.
In turn, this can be coupled to an action, like typing the inteded letter A or B
in Figure~\ref{fig:bci/loop}.
The \ac{bci} paradigm comprises the manner of stimulation (if applicable) and the
exact task performed by the user, and is often closely linked to the manner of
feedback in the case of a closed-loop system.
As shown in Section~\ref{sec:bci/paradigms}, there is a wide variety of \ac{bci}
paradigms based on different systems in the brain.

@@

The brain's activity can be recorded using various neuroimaging
technologies.
These can range from brain scans~\cite{Weiskopf2004} using functional magnetic resonance imaging
(fMRI) to more portable technologies like acoustic signals obtained by functional ultrasound imaging
(fUS)~\cite{Zheng2023} and near-infrared spectroscopy
(fNIRS)~\cite{Borgheai2020}.
The technologies above all measure the hemodynamic signal, which only indirectly
reflects brain activity.
Because of the desirable high temporal resolution~\cite{Easttom2021} and
practicality of electrophysiology, we will focus only on methods to record
electromagnetical activity originating from neuronal action potentials and
postsynaptic potentials.
The magnetical field of the brain can be measured using magnetoencephalography
(MEG)~\cite{Mellinger2007}.
While MEG is non-invasive and results in high-quality signals, the necessary
equipment is rather expensive and impractical.
Recent advances have been made using optically pumped magnetometers
(OPM-MEG)~\cite{Wittevrongel2021}, but this technology still falters outside of
lab conditions.

As a consequence, the \ac{bci} field relies heavily on the
electrical activity of the brain.
The \emph{invasiveness} of the technology forms an
important concept in determining its suitability for a given application and
patient.
As illustrated in Figure~\ref{fig:bci/recording}, invasive electrodes can often measure a very specific brain region and can result in better signal quality, at the cost of the risks
involved with brain surgery.

@@

\section{Paradigms}
\label{sec:bci/paradigms}
\todohsm{bring more structure, more subtitles}

@@

Paradigms that require little active participation of the user like
redirecting attention or initiating imagined actions are classified as \emph{passive
\ac{bci}}.
From the user's point of view, this concept is probably the most attractive,
since their intentions or cognitive state state would be inferred from just
their brain activity.\todohsm{in "passive bci", the user is not supposed to express "intentions". ragher start with active "}
However, without active user participation, it can be hard to establish reliable
communication or control.
Therefore, passive \ac{bci} are currently more suited to tasks like
emotion or affect monitoring~\cite{Torres2020,Libert2019}, workload
monitoring~\cite{Zanetti2021} or, more dependent on stimulation, error
detection~\cite{SiMohammed2020}.\todo{mention neurofeedback}
Furthermore, active paradigms\todo{"active" is used in a fuzzy way: active BCI is self-paced, reactive is in response to a cued event. Further down that becomes more clear.
The "active participation": is needed in both cases as shown in Fig. 1.3.
I suggest to say active participation} make it easier to quickly collect training data for
supervized machine learning since the conditions can be more easily controlled
by the user or the stimulation.

Paradigms with high active participation can be split up in \emph{active
\ac{bci}} and \emph{reactive \ac{bci}}.
Active \ac{bci} paradigms encode endogenous activity initiated by the user,
such as imagined movement or imagined speech.
These tasks can encode complex information, but current non-invasive
communication methods often limit the considered domain to a few movement
directions to control a cursor, or a few words~\cite{Panachakel2021}.
Invasive recordings are necessary for decoding more complex encoded
like natural speech~\cite{Metzger2023} or meaningful motor trajectories, like handwritten
symbols~\cite{Willett2021} or sign language~\cite{Branco2017}.
Active paradigms are also subject to large inter-subject variability due to the
complexity of the performed tasks.
Imagined speech or movements can be performed in a multitude of ways, which
can each have different neural representations for separate individuals.
This can make it hard to adapt the \ac{bci} to specific individuals, causing
poor performance and giving rise to the concept of \ac{bci}
illiteracy~\cite{Allison2010}\footnote{%
Recently, the concept of BCI illiteracy is increasingly being seen as outdated.
Instead of attributing poor performance to the user's inherent limitations,
critiques point out that the issue lies more with the design of the
systems themselves. BCI systems often fail to account for individual
variability in brain anatomy and cognitive strategies. As a result, what has
been labeled as "illiteracy" may, in fact, be a reflection of systems that are
not flexible enough to adapt to different users. This perspective
shifts the responsibility to the individual, rather than recognizing that
better, more adaptive systems could overcome many of these performance
challenges, thus standing in the way of progress.
Moreover, the term ``illiteracy'' itself carries negative
connotations, implying a user deficiency that reinforces normative assumptions,
when in reality, a more inclusive approach to design could enhance BCI
accessibility for a wider range of individuals. This evolving view calls for a
reframing of the problem, focusing on improving the adaptability of BCIs rather
than labeling users as illiterate~\cite{Becker2022,Thompson2019}.
}.
\todo{user training more than with reactive, task instruction}

Reactive \ac{bci} take another approach, by providing a discrete set of
sensory stimulations towards which the user's attention can directed.
Compared to active \ac{bci}, reactive paradigms usually require more effort\todohsm{is this the case?}
and can induce fatigue due to the constant stimulation.
These paradigms are also somewhat less intuitive compared to e.g.\ speech, and
their expressive power is limited by how many different stimuli can effectively
presented and attended within a reasonable timeframe.
Nevertheless, reactive \ac{bci} work well with \ac{eeg} recordings, and, most
importantly, they work for most people~\cite{Allison2010a,Edlinger2014}.

@@

\section{A case study: the visual oddball \acs{bci}}

### 2024-10-02 chapters/conclusion.tex
Comments hakim

\section{General discussion}

\subsection{Obtained results}

\subsubsection{\ac{erp} decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

@@

\subsubsection{Gaze-independent \ac{bci} healthy controls}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect\todomvh{not the right word, maybe the effect of covert VSA on
ERP...} observed for covert VSA also holds for split VSA.

@@

\subsection{Patient case studies}

@@

@@

\section{Current \& future work}
multicomponent alignment approach
hankel tensors to cope with jitter
select best classifier using eye tracking

\section{Recommendations}
\subsection{Working with patients}
immediately move to patients and on-line implementation, to be able to report
interpretable metrics
permanent working collaboration as a lab

start from patients
use all available signals
\subsection{Linear decoding methods}

>>>>

\section{Contributions}

\todo{come back on hypothesis}

\subsection{Developed \Ac{erp} decoders}
We introduced a covariance estimator using adaptive shrinkage (STBF-shrunk) and an estimator
exploiting prior knowledge about the spatiotemporal nature of the EEG signal
(STBF-struct).
We compared these estimators with the original formulation of the
spatiotemporal (STBF-emp)
beamformer and a state-of-the-art Riemannian Geometry method (XDAWN+RG) in an off-line P3 detection task on
an existing dataset.
Our results show that the structured estimator results in higher accuracy in
general, and specifically when training data are sparsely available.
Results can be computed faster and with
substantially less memory usage.
Since these algorithms are not paradigm-specific, the conclusions can be
generalized to other ERP-based BCI settings.
These results have been published in~\cite{VanDenKerchove2022}.

@@

\subsection{Gathered datasets}
The CVSA-ERP dataset consists of recordings of 15 healthy participants, mean age
$26.38\pm3.15$ years.
This dataset gives us insight in the ERP dynamics in overt, covert and the
novel split VSA condition, and confirms our hypothesis that P3 jitter has a
significant impact on performance in covert and split VSA, as well as that the
amplitude effect\todomvh{not the right word, maybe the effect of covert VSA on
ERP...} observed for covert VSA also holds for split VSA.

\todo{describe patients dataset}

\subsection{Investigated gaze-independence visual \acsp{bci}}

@@

@@

hard to evaluate experience in long term home use
unify clinical side and bci development side

@@

\section{Recommendations}
\subsection{Working with patients}
gaze independent visual, can be good current existing solution
take into accounrecommendations of \textcite{FriedOken2020}

immediately move to patients and on-line implementation, to be able to report
interpretable metrics
permanent working collaboration as a lab

In a mature field like visual bci, start from patients.
avoid situations where a problem is solved that doesn't pose itself in a
patient population
use all available capabilities and signals

%Have attention for the emotional impact

\subsection{Linear decoding methods}

\section{Current \& future work}
multicomponent alignment approach
hankel tensors to cope with jitter
select best classifier using eye tracking
expand on patient experiemnts

### 2024-10-02 chapters/gaze_independence.tex
Comments hakim

\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

@@

\section{Gaze-independent \ac{bci} paradigms}
\label{sec:gaze-independence}

Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and somatosensory
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and
user-dependent variability~\cite{Severens2014,Reichert2020b}.
\todo{insert table from riccio in text as numbers, not as table}

Secondly, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{Treder2010,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, these systems typically require users to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}

@@

Brain-Computer Interfaces (\ac{bci}s) decode brain activity with the aim to establish a direct communication
pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
devoid of these abilities, for whom \ac{bci}s can provide a means to communicate or to
control devices.
The quest for performant and affordable solutions is most evident in
visual \ac{bci}s based on the electroencephalogram (EEG), where it has led to a gamut of increasingly sophisticated decoders
and paradigms tailored to the needs of specific stimulation paradigms and
use contexts.
An effective and proven method is the visual event-related potential (\ac{erp})-based
oddball interface.
Here, targets are shown in short flashes on a computer screen.
Each time the user observes a target, an \ac{erp} is evoked, the presence of which can be
detected in the EEG-signal.
The \ac{erp} consists of multiple components, of which some are modulated by the
attention of the user, specifically the N2 and the P3 component.
Decoding this modulation allows for information transfer controlled by the
user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.
However, visual \ac{bci}s cannot operate efficiently when the user does not direct
their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
\todomvh{I am wondering whether goal statements and patient target groups
should be addressed in chapter 1.}

This leads to the often unadressed gaze-dependence paradox: the \ac{bci} target
population consists of patients who suffer from various degrees
of paralysis, or even from Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While \ac{bci}s are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking\todomvh{but with eye
control while the previous sentence points are absence of it}., studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Different degrees of eye motor impairment can render visual \ac{bci}s
uncomfortable or outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the \ac{bci}.
This points to the need for gaze-independent \ac{bci}s.

@@

This leads us to adopt a specific approach: improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy.
Hence, our working hypothesis is as follows:
\begin{quote}
	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
		can achieve a similar transfer rate in patient
		populations with eye motor deficits than other gaze-independent alternatives
    described in literature.}
\end{quote}

@@

\subsection{Interface design}

>>>>

\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}
\todohsm{move introduction of paper to paper}

@@

\section{Gaze-independent \ac{bci} paradigms}
\label{sec:gaze-independence}
\todohsm{rename to state of the art in gaze-independent bci. Address more systematically.}

Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and
somatosensory\todo{cite some}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and
user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

\todo{insert table from riccio in text as numbers, not as table}

This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the RSVP
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, these systems typically require users to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}

@@

\todohsm{ move next two paragraphs to introduction, more context than objectives}
Brain-Computer Interfaces (\ac{bci}s) decode brain activity with the aim to establish a direct communication
pathway bypassing speech and other forms of muscular activity. They have raised great hopes for patients
devoid of these abilities, for whom \ac{bci}s can provide a means to communicate or to
control devices.
An effective and proven method is the \ac{eeg}-based visual event-related potential (\ac{erp})-based
oddball interface~\cite{Wolpaw2018,Severens2020}.
Here, targets are shown in short flashes on a computer screen.
Each time the user observes a target, an \ac{erp} is evoked, the presence of which can be
detected in the EEG-signal.
The \ac{erp} consists of multiple components, of which some are modulated by the
attention of the user, specifically the N2 and the P3 component.
Decoding this modulation allows for information transfer controlled by the
user's brain activity, as sketched in Figure~\ref{fig:bci/loop}.
However, visual \ac{bci}s cannot operate efficiently when the user does not direct
their gaze onto the desired target~\cite{Brunner2010, Frenzel2011}.
\todomvh{I am wondering whether goal statements and patient target groups
should be addressed in chapter 1.}

This leads to the often underadressed gaze-dependence paradox: the \ac{bci} target
population consists of patients who suffer from various degrees
of paralysis, or even from Locked-in Syndrome (LiS), the complete loss of muscle
control with preserved consciousness.
While \ac{bci}s are most attractive as a solution for these patients
compared to other assistive technologies like eye-tracking\todomvh{but with eye
control while the previous sentence points are absence of it}., studies often
report bad performance in patient populations precisely due to the lack of adequate eye motor
control, or patients with gaze impairments are excluded.
Different degrees of eye motor impairment can render visual \ac{bci}s
uncomfortable or outright impossible to use.
When operating a visual \ac{bci}, the usual rapid series of forced saccades followed
by fixation is tiring over time, even for healthy control subjects.
A suitable alternative would allow the user to keep their eyes in an at-rest
position of their choice while operating the \ac{bci}.
This points to the need for gaze-independent \ac{bci}s.

@@

This leads us to adopt a specific approach: improve the performance of a
spatially organized visual oddball \ac{erp}-based brain
computer interface by using a suited decoding strategy.
Hence, our working hypothesis is as follows:
\todo{rewrite, this is not what we tested in the patient paper after all}
\begin{quote}
	\textit{A spatially organized visual oddball \ac{erp} \ac{bci} with a suited decoder
		can achieve a similar transfer rate in patient
		populations with eye motor deficits than other gaze-independent alternatives
    described in literature.}
\end{quote}

@@

\subsection{Interface design}
\todohsm{rename data collection. start with decoder design, second section data collection with experiments}
